Tracing the Hercules stream around the Galaxy

### 2010-05-19 tex/hercules.tex
groups-simulations: methodology section

\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath}
%\usepackage{color,hyperref}
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\figurenames}{\figurename s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\tvector}[1]{\boldsymbol{\vec{#1}}}
\newcommand{\vx}{\tvector{x}}
\newcommand{\vv}{\tvector{v}}
\newcommand{\Gaia}{\emph{Gaia}}
\newcommand{\apogee}{APOGEE}
\newcommand{\vR}{\ensuremath{v_R}}
\newcommand{\vphi}{\ensuremath{v_{\phi}}}
\newcommand{\vo}{\ensuremath{v_0}}
\newcommand{\Ro}{\ensuremath{R_0}}
\newcommand{\ro}{\Ro}

@@

We follow Dehnen's approach \citep{dehnen00a}. Only planar.

@@

\bibitem[Dehnen(1999a)]{dehnen99a}
  Dehnen,~W., 1999a, \aj, 118, 1190

\bibitem[Dehnen(1999b)]{dehnen99b}
  Dehnen,~W., 1999b, \aj, 118, 1201

\bibitem[Dehnen(1999c)]{dehnen99c}
  Dehnen,~W., 1999c, \apj, 524, L35

>>>>

\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath}
%\usepackage{color,hyperref}
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\figurenames}{\figurename s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\tvector}[1]{\boldsymbol{\vec{#1}}}
\newcommand{\vx}{\tvector{x}}
\newcommand{\vv}{\tvector{v}}
\newcommand{\Gaia}{\emph{Gaia}}
\newcommand{\apogee}{APOGEE}
\newcommand{\vR}{\ensuremath{v_R}}
\newcommand{\vphi}{\ensuremath{v_{\phi}}}
\newcommand{\vo}{\ensuremath{v_0}}
\newcommand{\Ro}{\ensuremath{R_0}}
\newcommand{\ro}{\Ro}
\newcommand{\Ab}{\ensuremath{A_b}}
\newcommand{\Rb}{\ensuremath{R_b}}
\newcommand{\Omegab}{\ensuremath{\Omega_b}}
\newcommand{\fdehnen}{\ensuremath{f_{\text{Dehnen}}}}
\newcommand{\sigmaR}{\ensuremath{\sigma_R}}
\newcommand{\rE}{\ensuremath{R_e}}
\newcommand{\Lc}{\ensuremath{L_c}}
\newcommand{\Rs}{\ensuremath{R_s}}
\newcommand{\Rsigma}{\ensuremath{R_{\sigma}}}
\newcommand{\Rolr}{\ensuremath{R_{\text{OLR}}}}

@@

To simulate the effect of the bar on stellar orbits near the Solar
circle, we follow the approach of \citet{dehnen00a}. For a fixed
position in the Galaxy, we evaluate the velocity distribution function
at values for the radial---toward the Galactic center---and
tangential---in the direction of Galactic rotation---velocity by
backward-integrating these velocities in the the Galactic potential to
obtain the intial orbit before bar formation. We posit that at this
epoch the old stellar disk was in a steady-state that can be described
by a simple distribution function that is a function of the energy and
angular momentum of the orbit alone. It follows from the collisionless
Boltzmann equation that the value of the velocity distribution at the
present position and velocity is equal to that of the steady-state
distribution function before bar formation evaluated at the initial
orbit.

Since the Hercules stream is only apparent in the planar motions in
the disk \citep[\eg][]{Bovy09a}, we only consider a two-dimensional
model for the Galaxy. We use a simple power-law rotation curve to
model the axisymmetric Galactic potential
\begin{equation}
v_c = \vo(R/\Ro)^\beta \,,
\end{equation}
and for most of the simulations below we specify this model further to
a flat rotation curve ($\beta = 0$). The model for the bar is given by
the following potential
\begin{equation}
\Phi_b(R,\phi) = \Ab(t) \cos [2(\phi - \Omegab t)] \times
\left\{ \begin{array}{ll} -(\Rb/R)^3\,, & \mathrm{for}\ R \geq
  \Rb\,,\\ (R/\Rb)^3-2\,, & \mathrm{for}\ R \leq \Rb\,. \end{array}
\right.
\end{equation}
Here \Omegab\ is the pattern speed of the bar and \Rb\ is the bar
radius, which is fixed to be 80\,percent of the bar's corotation
radius. The bar is grown smoothly during a time $t_1$, always set to
half of the total integration time in what follows, following the
prescription
\begin{equation}
\Ab(t) = \alpha \, \frac{\vo^2}{3}\,\left(\frac{\Ro}{\Rb}\right)^3
\times \left\{ \begin{array}{lll}
\frac{3}{16}\xi^5 - \frac{5}{8} \xi^3 + \frac{15}{16} \xi + \frac{1}{2}\,, & \xi \equiv 2\frac{t}{t_1}-1\,, & t \leq t_1\,,\\
1\,, & & t \geq t_1\,.\end{array} \right. 
\end{equation}

For the distribution function of the stellar disk before bar
formation, we use a Dehnen distribution function \citep{dehnen99b}
given by
\begin{equation}
\fdehnen(E,L) \propto \frac{\Sigma(\rE)}{\sigmaR^2(\rE)} \, \exp\left[ \frac{\Omega(\rE)\left[L-\Lc(E)\right]}{\sigmaR^2(\rE)}\right]\,,
\end{equation}
where \rE, \Lc, and $\Omega(\rE)$ are the radius, the angular
momentum, and the angular frequency, respectively, of the circular
orbit with energy $E$. Using the procedure given in \sectionname~3.2
of \citet{dehnen99b}, we choose the $\Sigma(R)$ and $\sigmaR$
functions such that they reproduce a disk with exponential surface
density and velocity dispersion profiles 
\begin{equation}
\Sigma(R) = \Sigma_0 \exp\left(-R/\Rs\right)\,, \qquad 
\sigmaR(R) = \sigma_0 \exp\left(-R/\Rsigma\right)\,,
\end{equation}
to an accuracy of a fraction of a percent at all radii. Our fiducial
model for the distribution function has \Rs = \Ro /3 and \Rsigma =
3\Rs, and $\sigma_0$ such that $\sigmaR(\Ro) = 0.2 \vo$.

As shown by \citet{dehnen99c,dehnen00a}, the model described above
that places the Sun at \Ro = 0.9\Rolr---where \Rolr\ is the radius at
which a circular orbit is in the outer Lindblad resonance with the
bar---at a current angle of 25$^{\circ}$ with the bar, with a bar
strength $\alpha = 0.01$, and integrating for four bar periods,
reproduces the Hercules feature in the local velocity distribution
while agreeing with other photometric and kinematical observations of
the bar. In what follows, unless indicated otherwise, we fix all of
the parameters of the model at these fiducial values.

@@

%\bibitem[Dehnen(1999a)]{dehnen99a}
%  Dehnen,~W., 1999a, \aj, 118, 1190

\bibitem[Dehnen(1999a)]{dehnen99b}
  Dehnen,~W., 1999a, \aj, 118, 1201

\bibitem[Dehnen(1999b)]{dehnen99c}
  Dehnen,~W., 1999b, \apj, 524, L35
